% © 2025 Daniel P. Madden — Custodial Author
% AI OSI Stack v5.0-open-core (Civic Standard Edition)

% © 2025 Daniel P. Madden. Custodial Edition – AI OSI Stack v5.0-open-core.
% Unauthorized reproductions or derivatives are not recognized custodial works.
% Refer to CANONICAL_PROVENANCE.yaml for official verification.
% SPDX-License-Identifier: CC-BY-SA-4.0

% AI OSI Stack v5 — Canonical Edition
% Appendix P — Advisory Integrity Workflow
\narrative{
Appendix P documents the advisory integrity workflow for AI OSI Stack v5. Instead of prescriptive signing ceremonies, custodians maintain
transparency by sharing the recommended steps for independent checksum capture. The manifest in \texttt{meta/v5-manifest.yaml} and the
Integrity Notice describe how to reproduce the canonical PDF and accompanying artefacts while acknowledging the absence of authoritative
cryptographic proofs. This appendix keeps those instructions inside the canon so readers understand how provenance is maintained in practice.
}

\narrative{
Integrity steward Imani walks practitioners through the advisory steps:
\begin{enumerate}
  \item Collect the canonical artefacts (PDF, manifest, AEIP schemas) from the repository or a trusted mirror.
  \item Generate local checksums using tooling such as \texttt{sha512sum} or \texttt{shasum -a 512}.
  \item Record the values in a personal or institutional log alongside timestamps and the active commit hash.
  \item Preserve the log with the artefacts and Integrity Notice to support future audits or archival retrieval.
\end{enumerate}
The workflow emphasises reproducibility and traceability rather than centralized attestation.
}

\begin{lstlisting}[language=yaml,caption={Advisory checksum log template}]
artefacts:
  - name: ai-osi-stack-v5.pdf
    path: versions/ai-osi-stack-v5.pdf
    sha512_recorded_by_verifier: "<your computed value>"
    recorded_at: "<ISO 8601 timestamp>"
    notes: "Captured by release engineer prior to archival upload."
  - name: manifest
    path: meta/v5-manifest.yaml
    sha512_recorded_by_verifier: "<your computed value>"
    recorded_at: "<ISO 8601 timestamp>"
    notes: "Stored with PDF and Integrity Notice."
\end{lstlisting}

\normative{
Custodians \shall{publish the advisory workflow} and \shall{encourage downstream users to maintain their own checksum logs}. Oversight bodies
\should{document the tools and timestamps used during verification}. Practitioners \may{supplement the workflow with additional controls}, but
they \shall{avoid representing the repository as cryptographically signed or formally certified}. Any detected divergence in recorded
checksums \shall{trigger a review of mirrors and storage locations before continued distribution}.
}