% AI OSI Stack v5 — Canonical Edition
% Appendix I — Security & Compliance Crosswalks
\narrative{
Appendix I maps the Stack’s safeguards to international security and compliance frameworks. Security architect Chen and compliance lead Aisha host the Crosswalk Atelier, where they align AI OSI controls with Annex IV, ISO/IEC 27001, NIST SP 800-53, GDPR, and sector-specific regulations. The narrative portrays crosswalking as a collaborative craft: spreadsheets become stories about harmonising obligations without diluting civic values.

Chen explains how Layer 3 model development controls correspond to secure development practices, while Aisha demonstrates how privacy provisions map to GDPR articles. They use AEIP manifests as connective tissue, ensuring that evidence collected for one framework satisfies others. When overlaps produce redundancy, the team designs integrated controls that respect the non-revocable principle. The appendix includes scenarios: a healthcare deployment aligning with HIPAA, a transportation system reconciling critical infrastructure standards, and an international partnership navigating divergent cybersecurity laws.

The narrative stresses adaptability. As regulations evolve, the Crosswalk Atelier reconvenes to update mappings, consult legal experts, and solicit civic feedback. Appendices E, K, and M inform risk prioritisation, ensuring human rights remain central. The appendix closes with guidance for organisations to localise crosswalks while maintaining fidelity to the canon.
}
\normative{
Security and compliance crosswalks \shall{} document control mappings between AI OSI obligations and applicable frameworks, including Annex IV, GDPR, ISO/IEC 27001, NIST SP 800-53, and relevant sector regulations. Each mapping MUST include control descriptions, responsible roles, evidence requirements, and AEIP manifest references.

Crosswalks \shall{} be reviewed at least semi-annually or upon regulatory change. Reviews MUST involve legal counsel, security engineers, privacy officers, and civic representatives. Updates \shall{} be recorded in Appendix C change logs and Appendix L interpretive notes, with public summaries provided through Appendix N transparency tiers.

Implementing organisations \shall{} publish crosswalk attestations that confirm alignment status, identify gaps, and outline remediation plans. Attestations MUST be signed by custodial quorums per Appendix H and made available to regulators, partners, and citizens. Failure to maintain current crosswalks \shall{} be treated as a governance deficiency subject to meta-audit scrutiny.
}
