% © 2025 Daniel P. Madden — Custodial Author
% AI OSI Stack v5.0-open-core (Civic Standard Edition)

% © 2025 Daniel P. Madden. Custodial Edition – AI OSI Stack v5.0-open-core.
% Unauthorized reproductions or derivatives are not recognized custodial works.
% Refer to CANONICAL_PROVENANCE.yaml for official verification.
% SPDX-License-Identifier: CC-BY-SA-4.0

% AI OSI Stack v5 — Canonical Edition
% Appendix B — Incident Lifecycle and Escalation (ILE)
\narrative{
Appendix B restores the Incident Lifecycle and Escalation sequence from version~4. Malik, the remediation coordinator, walks the
reader through the command floor as detection alerts arrive, verification teams assemble, escalation authorities convene, oversight
witnesses observe, and closure artefacts are published. The choreography is precise: each stage emits an AEIP record, carries a time
expectation, and references the custodial roles capable of invoking overrides. The appendix is designed as an executable playbook;
any jurisdiction can adopt the ILE and immediately map obligations to their command structure.
}

\narrative{
Detection begins when telemetry, civic reports, or custodial audits signal anomalous behaviour. Within 60~minutes, verification teams
pull the relevant AEIP manifests, confirm integrity, and classify severity. Escalation teams convene if the incident meets predefined
thresholds—public harm indicators, rights infringements, or mandate breaches. Oversight actors join within 24~hours, logging their
attendance in \texttt{schemas/aeip/incident-report-schema.json}. Closure occurs only after remediation actions are executed, evidence is
published, and civic communications acknowledge outcomes. Timelines are transparent and testable because each transition is linked to
an AEIP evidence field.
}

\begin{center}
\begin{tikzpicture}[node distance=2.6cm, every node/.style={rectangle, rounded corners, draw=black, thick, align=center, text width=4.3cm}]
  \node (detect) {\textbf{Detection}\\Trigger: AEIP telemetry, civic report, or audit flag\\Timing: within 15 min log entry\\AEIP: \texttt{incident.timeline.detected_at}};
  \node (verify) [right=of detect] {\textbf{Verification}\\Trigger: Duty officer assignment\\Timing: assessment within 60 min\\AEIP: \texttt{incident.timeline.verified_at}};
  \node (escalate) [right=of verify] {\textbf{Escalation}\\Trigger: Severity \textgreater{}= Level 3\\Timing: convene within 6 h\\AEIP: \texttt{incident.escalation.invoked_at}};
  \node (oversight) [right=of escalate] {\textbf{Oversight}\\Trigger: Civic/ombuds quorum\\Timing: observers join \textless{} 24 h\\AEIP: \texttt{incident.oversight.attendance[]}};
  \node (closure) [right=of oversight] {\textbf{Closure}\\Trigger: Remediation verified\\Timing: publish \textless{} 14 days\\AEIP: \texttt{incident.closure.attested_at}};
  \draw[->, line width=1pt] (detect) -- node[above]{AEIP Notification} (verify);
  \draw[->, line width=1pt] (verify) -- node[above]{Severity Packet} (escalate);
  \draw[->, line width=1pt] (escalate) -- node[above]{Override Authority Memo} (oversight);
  \draw[->, line width=1pt] (oversight) -- node[above]{Public Briefing + AEIP bundle} (closure);
\end{tikzpicture}
\end{center}

\normative{
Custodians \shall{log detection events} in \texttt{schemas/aeip/incident-report-schema.json} within 15~minutes of confirmation. Detection
logs \shall{include source system, mandate reference, affected population, and immediate safeguards}. Duty officers \shall{notify verification
teams} through \texttt{schemas/aeip/instruction-log-schema.json} entries so every investigative command is auditable. Civic reporters \may{receive
automated acknowledgement receipts} referencing the incident identifier and expected verification window.
}

\normative{
Verification leads \shall{complete triage within 60~minutes}. They \shall{capture severity ratings, impacted layers, and provisional
harm assessments} inside \texttt{incident.assessment}. If severity reaches Level~3 or higher, escalation coordinators \shall{issue an Override
Authority notice} using \texttt{schemas/decision-rationale-record.jsonld}. Notices \shall{enumerate which custodians may suspend services and
what rollback procedures apply}. Failure to escalate within the window \shall{trigger an automated reminder and report to Appendix~H
custodial succession forums}.
}

\normative{
Oversight participation \shall{commence within 24~hours}. Ombuds, community witnesses, and regulators \shall{register attendance and
conflict-of-interest attestations} through \texttt{incident.oversight.attendance}. Oversight chairs \should{issue an interim transparency
statement} mapped to Appendix~K tiers, clarifying what information \may{be disclosed immediately}. All oversight deliberations \shall{produce
AEIP witness notes} linked to Appendix~L hermeneutic ledgers.
}

\normative{
Closure \shall{occur within 14~days} unless remediation complexity is documented and approved by oversight participants. Closure packets
\shall{include remediation steps, restitution offers, model updates, and publication references}. Custodians \shall{update Layer~7 governance
notices and Layer~6 deployment checklists} with the resolution details. Incidents \may{remain open beyond 14~days} only if oversight actors
\shall{publish a delay justification} with revised milestones and evidence requirements.
}

\normative{
All incident records \shall{feed Appendix~C maturity tracking}. Institutions \shall{publish quarterly ILE performance metrics}, including detection
mean time, verification completion rate, escalation compliance, and closure punctuality. Metrics \should{be included in Appendix~N attestations}
and \may{be embedded in Layer~0 mandate renewals}. Persistent non-compliance \shall{escalate to Appendix~O provenance reviews} and \shall{trigger
public accountability hearings} in Layer~8 participation forums.
}