% SPDX-License-Identifier: CC-BY-SA-4.0

% AI OSI Stack v5 — Canonical Edition
% Appendix K — Transparency & Participation Interface
\narrative{
Appendix K reunifies the tiered transparency model of version~5 with the escalation and reporting duties from version~4. Transparency
steward Rina co-leads a briefing with Layer~7 publishers and Layer~8 participation councils. Together they map every incident stage from
Appendix~B onto four disclosure tiers: Civic Overview, Community Insight, Technical Detail, and Secure Oversight. Each tier lists who must
receive information, which AEIP artefacts are shared, and how feedback loops tie into participation ledgers. The appendix reads like a
civic-technical contract: disclosures are triggered by events, verified by evidence, and measured against participation outcomes.
}

\narrative{
When an escalation is invoked, Rina updates the Civic Overview within one hour, publishing mandate context and safety posture. Community
Insight briefings follow within six hours, highlighting local impact, translation requirements, and restitution options. Technical Detail
packets align with Layer~7 Governance Publication obligations, distributing hashes, model updates, and remediation tasks. Secure Oversight
artefacts reach regulators and ombuds within 24~hours through encrypted channels, referencing the AEIP incident bundle and override
notices. Participation councils in Layer~8 confirm that disclosures answered civic questions; outstanding items become agenda entries for
open forums. Transparency thus becomes a feedback engine rather than a one-way broadcast.
}

\normative{
Transparency programmes \shall{maintain the four-tier matrix} with explicit mappings to Layer~7 and Layer~8 obligations. Each tier \shall{define
audience, evidence bundle, disclosure cadence, escalation trigger, and participation feedback path}. Civic Overview releases \shall{publish
plain-language summaries, rights reminders, and AEIP manifest links}. Community Insight updates \shall{document local impact, translation
coverage, and contact channels for appeals}. Technical Detail packets \shall{include checksum references, remediation task lists, and
links to \texttt{schemas/governance-decision-summary.jsonld}}. Secure Oversight transmissions \shall{contain encrypted incident dossiers,
audit hooks, and override authority memos}. No tier \may{omit evidence references or participation contact details}.
}

\normative{
During incidents, tier updates \shall{adhere to Appendix~B timing}. Civic Overview \shall{activate within one hour of detection}, Community
Insight \shall{follow within six hours}, Technical Detail \shall{publish within 12~hours or upon verification}, and Secure Oversight \shall{convene within 24~hours}. Missed windows \shall{trigger automated alerts to Appendix~H custodianship forums}. Oversight chairs \should{issue
an interim transparency bulletin} logging which tiers were served and which remain pending. Participation leads \may{extend disclosure windows}
only when oversight actors \shall{publish a justification with revised deadlines and AEIP evidence references}.
}

\normative{
Feedback captured through any tier \shall{be recorded in Appendix~L hermeneutic ledgers and Appendix~C change logs} when it alters policy or
implementation. Layer~8 councils \shall{publish participation minutes summarising how disclosures addressed community questions}. If
disclosures reveal new harms, custodians \shall{open Appendix~B incident workflows}. Transparency metrics—tier activation time, feedback
resolution rate, and participation satisfaction scores—\shall{be published quarterly in Appendix~N}. Institutions \should{expose dashboards}
showing live compliance status, and civic observers \may{request audits} when metrics fall below mandate thresholds.
}
