% SPDX-License-Identifier: CC-BY-SA-4.0

% AI OSI Stack v5 — Canonical Edition
% Appendix L — Hermeneutic Ledger & Interpretive Records
\narrative{
Appendix L reveals the hermeneutic backbone of the Stack. The narrative follows interpretive archivist Nyasha as she curates the Hermeneutic Ledger—a repository of deliberations, contextual analyses, and cultural narratives that explain why decisions were made. The ledger blends transcripts, essays, multimedia artefacts, and civic testimony. Nyasha emphasises that interpretive records are not ancillary; they are essential to translating technical actions into shared understanding.

The story traces a dispute resolution case. Citizens challenged an algorithmic decision affecting housing allocations. Nyasha documents the interpretive journey: community forums, expert panels, historical research, and the eventual policy revision. Each step is captured in the ledger with metadata linking to AEIP manifests, persona obligations, and appendices. The narrative illustrates how interpretive records ensure that future custodians comprehend the reasoning, not just the outcome.

The appendix also showcases creative contributions. Poets, musicians, and visual artists interpret the Stack’s principles, adding emotive layers that inform governance. Nyasha curates these pieces alongside formal analyses, demonstrating the ledger’s pluralism. Interpretive stewardship involves continuous maintenance—indexing records, ensuring accessibility, and inviting public annotations. The appendix concludes with a call for participation: everyone is encouraged to contribute interpretations, critiques, and lived experiences.
}
\normative{
The Hermeneutic Ledger \shall{} capture interpretive artefacts associated with major decisions, incidents, policy changes, and civic engagements. Entries MUST include narrative context, contributors, references to relevant AEIP manifests and appendices, and accessibility metadata. Ledger maintenance \shall{} ensure multilingual support, disability accommodations, and open licensing consistent with the canon’s terms.

Interpretive reviews \shall{} occur quarterly, evaluating whether records remain comprehensive, balanced, and reflective of affected communities. Reviews MUST involve custodians, community representatives, and subject-matter experts. Identified gaps \shall{} prompt targeted outreach and supplemental documentation.

Public contributions to the ledger \shall{} be acknowledged, curated, and moderated according to transparent guidelines. Decisions to accept, edit, or reject submissions MUST be recorded with rationale. The ledger \shall{} integrate with Appendix N attestation systems and Appendix C change logs, ensuring interpretive context accompanies formal updates. Preservation plans \shall{} include redundant storage and integrity checks aligned with Appendices H and O.
}
