% SPDX-License-Identifier: CC-BY-SA-4.0

% AI OSI Stack v5 — Canonical Edition
% Appendix C — Governance Transport and Maturity Ladders
\narrative{
Appendix C translates the Stack’s maturity expectations into a transportable protocol. Where prior editions emphasised historical
lineage, version~5 couples that heritage with the ladder used to ratify civic deployments: Emergent \textrightarrow{} Structured
\textrightarrow{} Integrated \textrightarrow{} Institutionalized. Each rung measures how obligations, evidence, and community
custodianship travel between institutions. The appendix explains how Governance Integration Index scores attach to layers,
mirroring AEIP v1.3 manifest fields so auditors can verify readiness without guesswork.
}

\narrative{
The ladder originates in v4’s practitioner toolkit. This edition restores it by binding every checkpoint to AEIP evidence. Readers
see how ratification bundles capture quorum, how transport packages document cross-jurisdiction reciprocity, and how institutional
embeddedness depends on statutory budgeting. The generated tables in this appendix are produced by
\texttt{ops/scripts/generate-tex-tables.py}; they remain authoritative whenever custodians regenerate the edition. Governance teams can
therefore compare their posture against the canonical ladder and publish deltas as part of Appendix~N public attestations.
}

\normative{
Governance stewards \shall{calibrate each deployment against the maturity ladder} prior to ratification. Emergent adoption \shall{record
quorum proofs and incident drills} in AEIP civic charter and incident schemas. Structured adoption \shall{evidence control-table
generation events and publication windows}, drawing from manifest metadata. Integrated adoption \shall{demonstrate cross-jurisdiction
signature reciprocity and participatory response timeliness}. Institutionalized adoption \shall{document statutory enshrinement and
independent audit attestations}. Deviations \shall{be published as governance debt in Appendix~L hermeneutic ledgers} until resolved.
}

\normative{
When transporting the Stack between custodial homes, transition leads \shall{use the Governance Transport Grid} to certify which
layers are delegation-ready. Each row in the grid \shall{map to AEIP transport packages} whose checksums appear in the Integrity
Manifest appendix. Receiving institutions \should{acknowledge outstanding criteria and publish remediation timelines} before
assuming control. Any attempt to transfer responsibilities without satisfying the Structured stage for layers~0–4 \shall{trigger an
escalation to Appendix~B incident workflows} and \shall{pause deployment until governance debt is cleared}.
}

\begin{quote}
\textbf{Practitioner Note.} The data driving these tables lives in \texttt{meta/v5-maturity.json}. Regenerating this appendix with
\texttt{ops/scripts/generate-tex-tables.py} keeps narrative claims aligned with evidence structures.\end{quote}

\input{appendices/generated/appendix-c-maturity-tables}

\normative{
Custodians \shall{update the Governance Integration Index} whenever maturity checkpoints shift due to regulation, treaty, or
community mandate. Updates \shall{include AEIP manifest references and signature bundles} appended to Appendix~O. Oversight actors
\should{audit maturity declarations annually} and \may{issue public advisories} when institutions lag behind expected stages.
}
