% SPDX-License-Identifier: CC-BY-NC-ND-4.0
\chapter{Appendix D — Audit Structure}
\section*{Narrative Intent}
This appendix defines the structure and metadata required for audits referenced across Layers 03, 11,
and 12. It ensures audit artifacts remain comparable, reproducible, and publicly reviewable.

\section*{Normative Clauses}
\begin{enumerate}
  \item \textbf{Template Stewardship.} Custodians SHALL maintain standard audit templates including scope,
        methods, findings, and remediation commitments.
  \item \textbf{Archival Integrity.} Completed audits SHALL be archived with immutable identifiers and linked to
        related governance publications.
  \item \textbf{Public Summaries.} For audits affecting civic services, a plain-language summary SHALL be
        published alongside the technical report.
\end{enumerate}

\section*{Audit Metadata Fields}
\begin{description}
  \item[Scope ID] Cross-reference to layer or appendix under review.
  \item[Methodology] Tests, benchmarks, or interviews performed.
  \item[Findings] Confirmed strengths, gaps, and associated risk levels.
  \item[Remediation Plan] Actions, owners, and due dates linked to Appendix C change logs.
\end{description}

\section*{Plain-Speak Summary}
Audits follow a shared structure so reviewers can compare results. Each audit spells out what was
checked, what was found, and who is fixing any issues.

\cleardoublepage

% Authored and maintained solely by the Custodial Editorial Committee.
% This is a non-operational, publication-grade governance artifact.
% No AEIP runtime specs or machine-readable schemas are included.
