% AI OSI Stack v5 — Canonical Edition
% Appendix C — Change Log & Lineage
\narrative{
Appendix C tells the story of evolution. It guides readers through the lineage of the Stack, documenting how each version emerged from debates, experiments, and collective learning. The narrative unfolds as a guided tour of the Canonical Lineage Gallery, where timelines, ledger excerpts, and oral histories are displayed side by side. A curator named Elena narrates how v1 focused on establishing civic mandate, how v3 introduced federated governance pilots, and how v5 consolidates appendices, ledgers, and AEIP integration into a cohesive canon.

Visitors witness pivotal moments. The gallery showcases annotated AEIP manifests from crisis responses, persona revisions that reshaped service design, and international endorsements that expanded the Stack’s jurisdictional reach. Elena emphasises that lineage is not merely chronology but interpretation. Each exhibit includes commentary from citizens who experienced the policies, engineers who implemented controls, and custodians who navigated legal frameworks. The hermeneutic ledger provides audio transcripts of deliberations, preserving nuance that static documents cannot.

The appendix also recounts how the non-revocable principle was reaffirmed. After a contentious debate about surveillance technologies, the Stack codified safeguards that centre consent and proportionality. That decision is presented as a lineage milestone, linking Appendices E, K, and O. The narrative concludes by inviting readers to contribute to future lineage entries through public attestation channels, reinforcing that the canon’s history is co-authored by the community.
}
\normative{
All canonical changes SHALL be recorded in the change log with version identifiers, effective dates, responsible custodians, and references to affected chapters or appendices. Entries MUST include narrative summaries, normative rationale, and AEIP manifest identifiers that capture evidence of deliberation and approval. Change records SHALL be cross-linked to Appendix L interpretive notes and Appendix O provenance signatures.

Lineage updates SHALL be published in accessible formats, including interactive timelines and machine-readable datasets. Any alteration to canonical principles, governance structures, or obligations MUST undergo public consultation via Appendix N channels and receive custodial quorum approval per Appendix H. Rejected proposals SHALL also be logged, with rationale preserved for future reference.

Annual lineage reviews SHALL evaluate whether change logs remain comprehensive, accurate, and comprehensible to civic audiences. Reviews MUST document gaps, corrective actions, and recommendations for archive enhancement. Preservation plans SHALL ensure redundant storage of lineage records across custodial jurisdictions, safeguarding them against tampering or loss.
}
