% SPDX-License-Identifier: CC-BY-NC-ND-4.0
\chapter{Layer 00 — Civic Mandate}
\section{Narrative Intent}
\subsection*{Overview}
Layer 00 establishes the civic obligations that authorize the remainder of the Stack. It reframes the
Version 4 civic charter into a layered mandate with explicit public stewardship checkpoints, anchoring
the specification in shared democratic accountability. The intent is to translate civic expectations into
operational guardrails that remain legible to policymakers, operators, and the communities they serve.

This layer also clarifies the custodial relationship between the specification stewards and the public.
It introduces structured feedback loops that allow civic stakeholders to surface concerns, participate in
governance reviews, and track how their input informs revisions to downstream layers.

\section{Normative Clauses}
\subsection*{Mandated Obligations}
\begin{enumerate}
  \item \textbf{Civic Accountability Charter.} The custodian SHALL publish and maintain a civic accountability
        charter that enumerates public stewardship checkpoints, including schedules for consultation,
        responses to public comment, and traceability to the v4→v5 crosswalk commitments.
  \item \textbf{Participation Pathways.} Participatory mechanisms, including open comment periods and civic
        audit invitations, SHALL be documented with onboarding guidance, accessibility accommodations,
        and transparent disposition tracking.
  \item \textbf{Alignment Verification.} Every downstream layer SHALL reference the civic accountability
        charter to demonstrate how its controls map to the mandate, ensuring that operational decisions
        remain grounded in democratically endorsed values.
  \item \textbf{Custodial Transparency.} The custodial editorial committee SHALL maintain a public log of
        mandate-related decisions, capturing rationale, data sources, and the steward responsible for
        execution to support independent oversight.
\end{enumerate}

\section{Plain-Speak Summary}
Layer 00 spells out who holds the responsibility to the public and how they prove it. The stewards commit
to open consultation, transparent decision logs, and clear links between civic promises and technical
requirements. Every other layer must show how it honors these promises so that the public can see how
its expectations are carried forward.

\emph{License: Creative Commons Attribution–NonCommercial–NoDerivatives 4.0 International.}

\cleardoublepage

% Authored and maintained solely by the Custodial Editorial Committee.
% This is a non-operational, publication-grade governance artifact.
% No AEIP runtime specs or machine-readable schemas are included.
