% AI OSI Stack v5 — Canonical Edition
% Daniel P. Madden — Independent AI Researcher
% Created: 2025-11-04
% File: 00_Introduction_and_Purpose.tex
% Purpose: Chapter 00 rewritten in Phase 3 dual-register style
\narrative{
Governance is treated here as architecture rather than afterthought. The AI OSI Stack opens with a recognition that artificial intelligence is already entangled with water, energy, transport, health, and justice systems that citizens depend on daily. The introduction frames Version 5 as a public good, co-designed with civic forums, regulators, engineers, and researchers who insisted that legitimacy precede capability. Each layer is therefore a load-bearing element: Layer 0 confers civic mandate, Layer 1 encodes ethical intent, Layer 2 preserves data dignity, Layers 3–5 operationalise development, instruction, and reasoning integrity, while Layers 6–8 (retained from prior work) publish governance actions and honour civic participation. By narrating governance as an architecture, the chapter guides readers through how mandates become infrastructure, how evidence becomes civic memory, and how custodianship sustains continuity across jurisdictions.

The problem statement is clear: without verifiable structures, AI deployments default to opaque systems that erode trust and concentrate power. Communities asked for a stack that could be inspected, contested, and renewed. Version 5 responds with AEIP lifecycle scaffolding, Annex crosswalks, and interpretive ledgers that translate social expectations into operational rituals. Governance is not described as compliance paperwork; it is the construction of relational accountability between those who build, those who deploy, and those who live with the outcomes. The introduction signals to all audiences—citizens, policymakers, implementers, legal advisors, academics, and custodians—that their roles are interdependent, and that transparency only protects the public when anchored in rights-preserving design.
}
\normative{
\begin{itemize}
  \item Adopting entities SHALL implement Layers 0–8 of the AI OSI Stack in their entirety, mapping obligations to operational controls and registering coverage through AEIP lifecycle records (AEIP §O.3).
  \item Custodians and operators SHALL maintain AEIP evidence bundles that capture mandates, data stewardship controls, model assurance logs, instruction governance decisions, and reasoning interface audits, ensuring integrity across updates (AEIP §O.5).
  \item Governance Disclosure Systems (GDS) SHALL be published and continuously updated with canonical hashes, risk statements, steward contacts, and change logs consistent with Appendix E §3 and Appendix I crosswalk requirements.
  \item Civic appeal and remediation pathways SHALL be supported, documented, and actively maintained so that citizens can contest decisions, request redress, and monitor corrective actions without sacrificing privacy or due process (Appendix H §2).
\end{itemize}
}
\clearpage
