% SPDX-License-Identifier: CC-BY-NC-ND-4.0
\chapter{Layer 1 — Ethical Charter}
\section{Narrative Intent}
Layer 1 formalizes the ethical frame that all custodial actions must satisfy. Building on the civic mandate,
it articulates the shared principles—dignity, justice, proportionality, and explainability—that temper the
Stack’s technical ambitions. The narrative intent is to ensure that ethical reflection precedes deployment
choices, informs lifecycle decisions, and supplies the language for meaningful civic dialogue.

This charter also operationalizes reciprocity between affected communities and custodial stewards.
It commits the specification to protecting individuals from harm, centering marginalized voices in
review cycles, and maintaining ethical debt registers when trade-offs are accepted.

\section{Normative Clauses}
\begin{enumerate}
  \item \textbf{Principle Declaration.} The custodian SHALL maintain a canonical statement of ethical principles
        aligned to the civic mandate and update it whenever public comment materially shifts ethical
        expectations.
  \item \textbf{Impact Anticipation.} Each downstream control SHALL include an ethical impact annotation that
        outlines anticipated benefits, potential harms, and the mitigation measures chosen by the custodial
        stewards.
  \item \textbf{Reciprocity Assurance.} Governance processes SHALL document how affected parties were informed,
        consulted, and, where feasible, empowered to shape the resulting decision.
  \item \textbf{Ethical Debt Register.} When trade-offs are accepted, the custodian SHALL log the deviation,
        justify the temporary accommodation, and schedule the review checkpoint at which the debt must be
        reconciled or renewed.
\end{enumerate}

\section{Plain-Speak Summary}
The ethical charter lays down the values that steer every later requirement. It says the team must publish
its moral commitments, think through impacts before acting, listen to people who carry the risk, and keep
records of any compromise so it can be fixed or retired. Ethics is treated as an ongoing practice, not a one-
time declaration.

\cleardoublepage

% Authored and maintained solely by the Custodial Editorial Committee.
% This is a non-operational, publication-grade governance artifact.
% No AEIP runtime specs or machine-readable schemas are included.
