% AI OSI Stack v5 — Canonical Edition
% Daniel P. Madden — Independent AI Researcher
% Created: 2025-11-04
% File: 02_Philosophical_Foundations.tex
% Purpose: Chapter 02 merging philosophical closure and Epistemology by Design
\narrative{
Philosophical foundations transform the stack from procedural governance into an intelligible moral system. Version 5 merges the earlier “Philosophical Closure” work with Epistemology by Design, treating ontology, epistemology, and axiology as auditable metadata rather than abstract essays. Ontology clarifies what entities exist in the governance model: citizens, custodians, data artefacts, models, personas, and civic ledgers. Epistemology describes how the stack knows what it knows: AEIP receipts, interpretive records, community testimonies, and cross-jurisdictional audits. Axiology anchors value commitments: human dignity, proportionality, restorative justice, and the non-revocable clause “Transparency must never become surveillance.” By cataloguing these assumptions, the stack ensures that each layer’s technical decisions remain accountable to declared values.

Narratively, this chapter shows how philosophical clarity prevents governance drift. When a deployment team faces ambiguous data provenance or contested model behaviour, they consult the ontological map to determine which actors hold legitimacy. When oversight committees ask how a recommendation was justified, they interrogate the epistemic chain recorded in AEIP Signalled Interpretive Records. When citizens challenge an outcome, the axiological commitments provide benchmarks for remediation. The philosophical metadata therefore operates as a civic compass, guiding iterative updates without erasing founding principles. It also facilitates federated collaboration: jurisdictions can compare their own assumptions with those declared here, aligning or documenting divergence in Appendix N and AEIP exchange bundles.
}
\normative{
\begin{itemize}
  \item All ontological, epistemic, and axiological assumptions informing AI OSI Stack implementations SHALL be explicitly declared, versioned, and published through AEIP Signalled Interpretive Records (SIRs) in accordance with (AEIP §O.5).
  \item Custodians SHALL review philosophical metadata during every major release or policy update, documenting continuity or change in Appendix E §3 crosswalks and the Canonical Provenance Statement.
  \item Implementers SHALL align deployment decisions with the declared axiological commitments; deviations MUST trigger remedial analysis, civic consultation where applicable, and publication of the decision rationale within seven days.
  \item The non-revocable clause “Transparency must never become surveillance.” SHALL remain encoded in all philosophical metadata and SHALL inform privacy-preserving design decisions, appeal processes, and publication practices.
\end{itemize}
}
\clearpage
