% SPDX-License-Identifier: CC-BY-NC-ND-4.0
\chapter{Layer 07 — Governance Publication}
\section{Narrative Intent}
\subsection*{Overview}
Layer 07 describes how governance artifacts are published, validated, and archived for public
consumption. It refines the publication cadence from Version 4, mandating open audit artifacts and
linking outputs to the civic accountability charter. The narrative underscores transparency as a
continuous obligation rather than an annual report.

\section{Normative Clauses}
\subsection*{Publication Duties}
\begin{enumerate}
  \item \textbf{Cadence Commitment.} Custodians SHALL publish governance updates on a predictable cadence,
        announcing timelines and deviation explanations.
  \item \textbf{Audit Artifact Release.} Audit summaries, methodologies, and remediation plans SHALL be made
        publicly accessible, with sensitive data redacted but obligations intact.
  \item \textbf{Evidence Registry Integration.} Published materials SHALL reference the epistemic evidence
        registry, enabling independent verification of claims.
  \item \textbf{Accessibility Standards.} All publications SHALL meet accessibility requirements and provide
        multi-language summaries where civic mandates require.
\end{enumerate}

\section{Plain-Speak Summary}
The public must know what is happening and when. This layer requires frequent, accessible updates that
include the evidence behind decisions. If dates change, the reason is documented. Transparency becomes
a routine habit, not a ceremonial gesture.

\emph{License: Creative Commons Attribution–NonCommercial–NoDerivatives 4.0 International.}

\cleardoublepage

% Authored and maintained solely by the Custodial Editorial Committee.
% This is a non-operational, publication-grade governance artifact.
% No AEIP runtime specs or machine-readable schemas are included.
