% SPDX-License-Identifier: CC-BY-NC-ND-4.0
\chapter{Layer 09 — Epistemic Infrastructure Protocol}
\section{Narrative Intent}
\subsection*{Overview}
Layer 09 introduces epistemic health indicators and public evidence registries. Its narrative intent is to
anchor every claim in verifiable artifacts, allowing civic reviewers to test whether the Stack’s decisions
are grounded in sound evidence and transparent reasoning pathways.

\section{Normative Clauses}
\subsection*{Epistemic Controls}
\begin{enumerate}
  \item \textbf{Evidence Registry.} Custodians SHALL maintain a public-facing registry of evidence packages,
        linking datasets, analyses, and attestations used to support governance decisions.
  \item \textbf{Health Indicators.} Epistemic health metrics—coverage, freshness, contradiction rates—SHALL be
        defined, monitored, and reported at each governance publication cycle.
  \item \textbf{Verification Hooks.} External reviewers SHALL be granted the ability to reproduce or challenge
        evidence packages through sandboxed verification environments.
  \item \textbf{Redaction Discipline.} When sensitive material is withheld, custodians SHALL document the legal
        or ethical basis and provide alternative summaries to sustain civic trust.
\end{enumerate}

\section{Plain-Speak Summary}
This layer keeps the knowledge base honest. Every important claim needs documented proof, and outsiders
get controlled ways to test that proof. If something cannot be shown, the team must explain why and offer
as much context as possible.

\emph{License: Creative Commons Attribution–NonCommercial–NoDerivatives 4.0 International.}

\cleardoublepage

% Authored and maintained solely by the Custodial Editorial Committee.
% This is a non-operational, publication-grade governance artifact.
% No AEIP runtime specs or machine-readable schemas are included.
