% SPDX-License-Identifier: CC-BY-NC-ND-4.0
\chapter{Layer 11 — Implementation Verification}
\section{Narrative Intent}
\subsection*{Overview}
Layer 11 tightens verification benchmarks and connects them to publicly accessible civic verification
tooling. It ensures every control has measurable tests, independent validation pathways, and documented
results that feed into governance publications.

\section{Normative Clauses}
\subsection*{Verification Duties}
\begin{enumerate}
  \item \textbf{Benchmark Definition.} Custodians SHALL maintain benchmark suites that evaluate compliance,
        safety, and civic impact across representative scenarios.
  \item \textbf{Independent Review.} Verification results SHALL undergo independent review, either by internal
        teams with separation of duties or by accredited third parties.
  \item \textbf{Public Traceability.} Verification artifacts SHALL be indexed in Appendix D so civic observers can
        confirm the provenance and status of each test.
  \item \textbf{Continuous Improvement.} Benchmark gaps and failures SHALL be logged with remediation plans, and
        progress SHALL be tracked in governance publications.
\end{enumerate}

\section{Plain-Speak Summary}
Verification shows whether promises were kept. This layer sets the expectations for building, running,
and sharing tests. If a test fails or does not exist, that gap is logged and fixed in full view of the public.

\emph{License: Creative Commons Attribution–NonCommercial–NoDerivatives 4.0 International.}

\cleardoublepage
