% SPDX-License-Identifier: CC-BY-NC-ND-4.0
\chapter{Layer 12 — Strategic Resilience \& Risk Mitigation}
\section{Narrative Intent}
\subsection*{Overview}
Layer 12 integrates resilience stress-testing and escalation playbooks. It translates strategic risk
insights into actionable plans that keep civic services available during crises. The narrative focuses on
anticipating disruption and demonstrating how the Stack adapts without abandoning civic commitments.

\section{Normative Clauses}
\subsection*{Resilience Controls}
\begin{enumerate}
  \item \textbf{Scenario Planning.} Custodians SHALL maintain multi-horizon risk scenarios, assessing likelihood,
        civic impact, and mitigation strategies.
  \item \textbf{Stress Testing.} Critical components SHALL undergo periodic stress tests, with findings logged and
        remediation tracked through Appendix C change logs.
  \item \textbf{Escalation Playbooks.} Incident response playbooks SHALL align with Appendix B procedures, defining
        roles, communication templates, and decision authorities.
  \item \textbf{Continuity Metrics.} Resilience performance metrics SHALL be reported to the public, including
        downtime, service degradation, and recovery times.
\end{enumerate}

\section{Plain-Speak Summary}
This layer prepares the Stack for rough conditions. Teams map out possible crises, rehearse their
responses, and share how well the system holds up. The public gets honest reports about disruptions and
how long recovery took.

\emph{License: Creative Commons Attribution–NonCommercial–NoDerivatives 4.0 International.}

\cleardoublepage

% Authored and maintained solely by the Custodial Editorial Committee.
% This is a non-operational, publication-grade governance artifact.
% No AEIP runtime specs or machine-readable schemas are included.
