% AI OSI Stack v5 — Canonical Edition
% Daniel P. Madden — Independent AI Researcher
% Created: 2025-11-04
% File: 16_Strategic_Resilience_and_Risk_Mitigation.tex
% Purpose: Canonical chapter content for strategic resilience
\narrative{
Strategic resilience is presented as both a moral covenant and a technical discipline. Civic custodians rehearse distributed incident playbooks that braid community welfare teams with engineering brigades, affirming that continuity of care is a shared obligation rather than an optional control. The stack curates scenario libraries drawn from plural epistemologies, using climate compacts, labor accords, and critical infrastructure registries to stress governance pathways before harm materialises. Resilience labs run red-blue adversarial simulations that confront cascading failure hypotheses, validating how fail-safe design, graceful degradation, and mutual aid contracts uphold the civic mandate under duress. (Appendix B §3–§5)

Narrative cases recount how hermeneutic sentinels track epistemic drift during crises, documenting contested facts and counter-speech in the ledger so that institutional memory remains trustworthy once the emergency has passed. Custodians deploy recovery swarms—rapid response cohorts empowered to re-establish data lineage, recalibrate risk tiers, and re-authorise affected services only after community advocates certify readiness. The narrative emphasises that each recovery cycle is archived as civic jurisprudence, folding lessons back into Update Plans and renewed solidarity accords. (Appendix I §4)
}
\normative{
Resilience governance SHALL maintain contingency AEIP packages for every critical asset class, including charter, compute, data, model, control, and deployment artefacts. Risk-tier thresholds MUST be codified in the ledger with explicit decision checkpoints, and any elevation beyond tolerated exposure SHALL trigger cross-functional escalation with citizen oversight present. Custodians SHALL operate audit loops that document rehearsal frequency, scenario variance, and ledger recovery proofs; results MUST be notarised under AEIP manifest tags referencing response posture and remediation status. (AEIP §O.6)

All recovery operations SHALL evidence fail-safe design by demonstrating redundancy in data custody, authority delegation, and communications. Lessons learned MUST be captured within Update Plan supplements and SHALL include commitments to repair epistemic injury for impacted communities. Non-compliance with resilience remits SHALL be logged as governance variance, reported to stewardship councils, and remediated through time-bound corrective plans that incorporate civic witness sign-off. (Appendix B §3–§5),(Appendix I §4)
}
