% AI OSI Stack v5 — Canonical Edition
% Daniel P. Madden — Independent AI Researcher
% Created: 2025-11-04
% File: 18_Interpretive_and_Temporal_Continuity.tex
% Purpose: Canonical chapter content for interpretive continuity
\narrative{
Interpretive and temporal continuity narrates how the stack preserves semantic intent across shifting eras, toolchains, and custodial cohorts. Storylines trace archival rituals in which community archivists, system librarians, and policy translators convene continuity councils to revoice canonical principles for contemporary contexts. These councils maintain cross-hash chains linking prior deliberations, decision rationales, and public commentaries so that future interpreters inherit the lineage of meaning rather than isolated excerpts. (Appendix L §3)

The narrative follows a temporal drift investigation where automated summarisation tools altered reparative justice language. Custodians replayed ledger diffs, consulted oral history stewards, and reconstructed the archival thread to restore the original commitments. Time capsules—annotated bundles of manifests, testimonies, and horizon scans—are sealed at each major release to guide successors through context-rich handovers, ensuring that the stack remains a living conversation rather than a static archive. (Appendix H §2)
}
\normative{
Interpretive continuity SHALL be proven through cross-hash chains that reference every canonical amendment, linking narrative and normative registers with their antecedents. Temporal stewardship MUST run ledger drift tests whenever toolchains, translation layers, or custodial teams change; results SHALL be logged with AEIP continuity tags and peer-reviewed by continuity councils. Any detected drift MUST trigger restorative documentation, including redlines, counter-narratives, and community ratification steps prior to re-publication. (AEIP §O.5–§O.7)

Successor custodians SHALL execute knowledge-transfer convenings that document interpretive briefs, horizon risks, and unresolved dissent. Archival migrations MUST preserve checksum integrity, semantic annotations, and contextual metadata, with public proofs of verification stored in the Hermeneutic Ledger. Where continuity breaks cannot be fully remedied, custodians SHALL issue variance statements detailing cause, community impact, and compensatory safeguards for future readers. (Appendix L §3),(Appendix H §2)
}
