% © 2025 Daniel P. Madden. Custodial Edition – AI OSI Stack v5.0-open-core.
% Unauthorized reproductions or derivatives are not recognized custodial works.
% Refer to CANONICAL_PROVENANCE.yaml for official verification.
% SPDX-License-Identifier: CC-BY-SA-4.0

% Implements: tools/control-tower
% AI OSI Stack v5 — Canonical Edition
% Daniel P. Madden — Independent AI Researcher
% Created: 2025-11-09
% File: 19_Epilogue_The_Stack_as_Living_Constitution.tex
% Purpose: Canonical chapter content for epilogue
\narrative{
The epilogue frames the AI OSI Stack as a living constitution—a covenant continually renewed through participation, accountability, and care. The narrative returns to the Civic Forum described in earlier chapters, now convened to celebrate the canonical release. Citizens, policymakers, engineers, and custodians gather to reflect on the journey from initial mandates to mature governance. The atmosphere is both ceremonial and practical: while gratitude is shared, the forum remains focused on sustaining obligations.

Speakers recount milestones. Community advocates describe how Layer 0 commitments empowered them to demand co-design. Engineers share stories of the first AEIP bundles that survived adversarial stress tests. International partners highlight moments when federated governance (Appendix G) allowed them to align local values with global standards. Each testimonial is paired with a projection of ledger entries, demonstrating that narrative memories align with verifiable records. The canon’s non-revocable principle is reiterated: transparency must never become surveillance, and vigilance is required to protect that balance.

The epilogue introduces future stewards. Youth delegates present the time capsule from Chapter 18, inviting the audience to imagine the Stack fifty years ahead. They ask: How will emerging technologies challenge existing safeguards? What new forms of solidarity will be necessary? Elders respond by offering lessons about humility, resilience, and the need to keep inviting dissent. The forum endorses a resolution committing to annual renewal assemblies where the canon is reviewed, amended, or reaffirmed through civic deliberation recorded in Appendix L.

As the gathering concludes, custodians seal the canonical release. They sign the master provenance statement (Appendix O), publish integrity hashes to the public ledger, and invite citizens to verify the PDF using Appendix N instructions. A moment of silence honours those who contributed labour, care, and critique. The epilogue emphasises that the Stack’s legitimacy depends on continued participation: every reader is asked to join as steward, witness, or challenger.

The final scene shifts to a quiet municipal library where a student accesses the Stack through the public archive. They cross-check the hash, read the appendices, and annotate questions for the next civic forum. The living constitution breathes through such acts of engagement. The canon remains open, accountable, and rooted in the shared conviction that transparency is a public good only when paired with dignity and restraint.
}

\subsection*{Triple Register}
\textbf{Narrative Intent:} The epilogue addresses the challenge of keeping the stack a living constitution, emphasising how stewardship, public oversight, and iterative amendments stay synchronised.
\textbf{Normative Clauses:}
\begin{itemize}
\item Custodians \shall{} log constitutional updates within \texttt{schemas/integrity-ledger-entry.jsonld} to evidence legitimacy.
\item Governance councils \shall{} publish amendment rationales via \texttt{schemas/governance-decision-summary.jsonld} for civic review.
\item Stewards \should{} refresh lifecycle commitments through \texttt{schemas/aeip-template.yaml} whenever the constitution evolves.
\end{itemize}
\textbf{Plain-Speak Summary:} The epilogue reminds readers that the stack must keep adapting without losing its soul. It records how changes are logged, explained, and adopted. People can see which commitments endure and which are reworked. The framework stays alive because every change is accountable.

\normative{
The canonical release \shall{} be accompanied by public verification materials, including integrity hashes, provenance signatures, and attestation guides aligned with Appendices N and O. Custodians MUST convene annual renewal assemblies to review canonical obligations, record interpretive updates, and invite civic participation. Outcomes \shall{} be documented in the hermeneutic ledger and incorporated into Appendix C change logs.

Citizens and partners \shall{} retain the right to petition for amendments, audits, or clarifications through processes defined in Appendices G, L, and N. Custodians MUST respond within defined timelines, providing evidence, rationale, and remedial actions where applicable. The Stack \shall{} maintain accessible archives, educational resources, and participation pathways to ensure that the canon functions as a living constitution rather than a static document.

Every distribution of the canonical edition \shall{} reference the non-revocable principle and the obligations to prevent transparency from morphing into surveillance. Implementers MUST honour persona commitments, epistemic safeguards, and human rights protections described throughout the canon. Failure to do so triggers custodial review, public notice, and potential succession under Appendix H, preserving the canon’s integrity for future generations.
}
\n\subsection*{Verification and Enforcement}
Conformance is evidenced through artefacts \texttt{schemas/integrity-ledger-entry.jsonld}, \texttt{schemas/governance-decision-summary.jsonld}, and \texttt{schemas/aeip-template.yaml} and corresponding AEIP audit records.