% © 2025 Daniel P. Madden — Custodial Author
% AI OSI Stack v5.0-open-core (Civic Standard Edition)

% © 2025 Daniel P. Madden. Custodial Edition – AI OSI Stack v5.0-open-core.
% Unauthorized reproductions or derivatives are not recognized custodial works.
% Refer to CANONICAL_PROVENANCE.yaml for official verification.
% SPDX-License-Identifier: CC-BY-SA-4.0

% Implements: tools/verify-aeip-signatures.py
% AI OSI Stack v5 — Canonical Edition
% Daniel P. Madden — Independent AI Researcher
% Created: 2025-11-09
% File: 12_AEIP_Framework_and_Operations.tex
% Purpose: Canonical chapter content for AEIP framework
\narrative{
Chapter 12 follows the AI Epistemic Infrastructure Protocol (AEIP) across an operational day. The story opens inside the Civic Evidence Exchange, a shared workspace where municipal archivists, system engineers, and independent auditors braid together receipts from Layers 0 through 11. AEIP v1.3 sits at the centre of the room as both schema and social contract: its JSON-LD manifests encode every design choice, mitigation, and consent that the Stack has promised to citizens. The narrative walks alongside a release historian named Noa who curates an integrity bundle for a new public-service recommender. Her task is to assemble prior commitments, fold in model validation artefacts, and publish a provenance signature that federated partners can trust without asking for privileged access.

Noa begins with source capture. AEIP collectors watch the Layer 3 model pipeline, Layer 5 reasoning exchange, and Layer 6 deployment rehearsals, emitting cryptographic hashes each time evidence crosses a lifecycle stage. The story details how privacy-aware descriptors differentiate among consent scopes, lawful bases, and restricted data classes so that transparency never mutates into surveillance. When a safety engineer attaches dark-launch telemetry, the AEIP validator notes the data minimisation posture and demands that pseudonymisation proofs accompany the metrics. Noa logs the conversation in the hermeneutic ledger, acknowledging that interpretive context is as valuable as binary pass/fail states.

The chapter shifts into the Federation Verification Lab, where partners from regional utilities and research consortia replay the bundle. AEIP’s operations are dramatized as a civic ritual: each partner brings their own validator, compares hash chains, and inspects cross references to Appendices F, I, and O.3–O.7. When a discrepancy arises—a mismatched timestamp due to daylight-saving transitions—the lab narrates the joint troubleshooting procedure. Instead of hiding the inconsistency, the team escalates to the Custodianship Desk, which issues an interpretive addendum explaining the time-source adjustment. The resolution reinforces the canon’s non-revocable principle: transparency retains fidelity because every deviation is explained, not concealed.

Lifecycle maintenance rounds out the narrative. AEIP operations staff convene weekly to review anomaly reports generated by automated monitors that watch for incomplete manifests, broken signature chains, or unreferenced assets. The monitors are personified as sentinels that never sleep yet remain subordinate to human custodians. Their alerts feed into the meta-audit queue described in Chapter 17, where auditors triage issues into remediation workflows. An incident drill illustrates how AEIP receipts help orchestrate rollback: when a federated partner simulates a hash collision attack, the operations team rotates keys, republishes evidence bundles, and anchors the change in the integrity ledger without disrupting public access. By sunset, Noa signs the updated bundle with a custodial quorum and publishes the ledger index that citizens can verify against Appendix N.

The narrative concludes with a reflection on memory. AEIP is portrayed as the Stack’s conscience, a protocol that remembers promises, surfaces rationales, and invites oversight. Its operations bind engineers, policymakers, and civic custodians together through shared observability. The final scene shows a student journalist verifying the bundle’s hash via the public attestation guide, demonstrating that AEIP’s machinery empowers civic literacy rather than overwhelming it.
}

\subsection*{Triple Register}
\textbf{Narrative Intent:} This chapter tackles the coordination burden of evidence gathering by explaining how the AEIP becomes the backbone for lifecycle assurance and multi-party accountability.
\textbf{Normative Clauses:}
\begin{itemize}
\item Implementers \shall{} instantiate assurance packages using \texttt{schemas/aeip-template.yaml} before work enters regulated stages.
\item Programme leads \shall{} align lifecycle checkpoints with the enumerations in \texttt{schemas/aeip-1-3.jsonld} to maintain interoperability across jurisdictions.
\item Custodians \should{} validate cross-layer linkages against \texttt{schemas/aeip/aeip-frame-schema.json} prior to closing audits.
\end{itemize}
\textbf{Plain-Speak Summary:} This chapter shows how the AEIP keeps every promise and artefact stitched together. It walks teams through the templates that must be filled out. Readers see how those records travel across legal and technical reviews. Following the AEIP makes audits faster and fairer.

\normative{
AEIP implementations \shall{} conform to schema version 1.3 and MUST declare artifact.type, lifecycle.stage, and provenance.signature fields for every manifest. Each manifest \shall{} reference the applicable layer path, Appendix cross-references, and AEIP §O.3–§O.7 clauses that justify retention, access, and disclosure. Evidence bundles MUST maintain an ordered chain of receipts using SHA-512 or stronger digests, with signing keys rotated according to Appendix H custodianship controls.

Operational tooling \shall{} enforce privacy.* validators that capture lawful basis, consent scope, and retention schedule before any manifest is accepted into the ledger. Telemetry or datasets lacking documented minimisation \shall{} be rejected until corrective actions restore compliance. Federation validators MUST support deterministic replays of manifest verification, including timezone normalisation, schema evolution notes, and interpretive addenda recorded within the hermeneutic ledger.

AEIP custodians \shall{} maintain weekly anomaly reviews that reconcile validator alerts, incomplete manifests, and signature expirations. Incident response involving evidence integrity MUST follow Appendix B remediation procedures, rotate keys, regenerate hashes, and append public notices within 24 hours. Public attestation endpoints \shall{} expose bundle indices, hash proofs, and verification instructions consistent with Appendix N so that citizens and partners can independently confirm authenticity without escalating surveillance.
}
\n\subsection*{Verification and Enforcement}
Conformance is evidenced through artefacts \texttt{schemas/aeip-template.yaml}, \texttt{schemas/aeip-1-3.jsonld}, and \texttt{schemas/aeip/aeip-frame-schema.json} and corresponding AEIP audit records.