
%---------------------------------------------------------
\chapter{Appendix J — EU Legal Alignment Tables}
%---------------------------------------------------------
% AI OSI Stack v5 — Canonical Edition
% Appendix J — EU Legal Alignment Tables
\narrative{
Appendix J demonstrates how the Stack aligns with European Union legal instruments, including the AI Act, GDPR, Digital Services Act (DSA), and fundamental rights charters. The narrative accompanies legal analyst Sofia as she compiles alignment tables in collaboration with municipal lawyers and civil-society monitors. They gather in the European Compliance Studio, a workspace filled with side-by-side translations, legal commentaries, and AEIP manifests.

Sofia walks readers through an AI Act conformity assessment: risk classification, Annex IV documentation, and post-market monitoring. The alignment table is narrated as a dialogue between legal text and civic obligations. Municipal lawyers contribute examples from public services, while monitors raise concerns about enforcement gaps. The story emphasises how AEIP manifests provide verifiable evidence to regulators, reducing administrative burden and enhancing trust.

The appendix also examines GDPR intersections. Data protection officers map lawful bases, consent flows, and data subject rights to Stack controls. They address complex scenarios such as joint controllership in federated deployments and cross-border data transfers. The narrative highlights the role of civic oversight committees in verifying compliance and advocating for additional safeguards when EU law leaves room for interpretation.

Sofia concludes by presenting dashboards that track legal alignment status across jurisdictions. The tables feed into Appendix I crosswalks and Appendix K transparency tiers, ensuring that citizens can understand how EU law shapes AI governance. The appendix underscores that legal alignment is iterative, requiring continuous dialogue between legislators, implementers, and the public.
}
\normative{
EU legal alignment tables SHALL document mappings between Stack obligations and relevant EU legal requirements, including AI Act articles, Annex IV criteria, GDPR provisions, DSA obligations, and Charter rights. Each table MUST list legal references, Stack controls, responsible roles, evidence artefacts, and compliance status. Tables SHALL be maintained for each jurisdiction deploying the Stack within the EU or engaging with EU residents.

Alignment reviews SHALL occur whenever EU regulations change, significant system updates occur, or annually—whichever happens first. Reviews MUST involve legal counsel, data protection officers, civic representatives, and custodians. Updates SHALL be recorded in Appendix C change logs and communicated through Appendix N transparency notices.

Non-compliance findings SHALL trigger remediation plans aligned with Appendix B procedures, including timelines, responsible owners, and verification checkpoints. Persistent misalignment SHALL escalate to federated governance forums (Appendix G) and may require suspension of affected services until compliance is restored.
}
%---------------------------------------------------------