%---------------------------------------------------------
\chapter{Appendix K — Civic Transparency Tiers And Contextual Disclosure}
%---------------------------------------------------------
% AI OSI Stack v5 — Canonical Edition
% Appendix K — Civic Transparency Tiers & Contextual Disclosure
\narrative{
Appendix K organises transparency into contextual tiers that respect privacy and security while empowering citizens. The narrative follows transparency steward Rina as she prepares disclosures for a new AI-enabled public service. She consults the Transparency Matrix, which defines four tiers: Civic Overview, Community Insight, Technical Detail, and Secure Oversight. Each tier targets specific audiences and includes guidance on format, frequency, and safeguards.

Rina begins with the Civic Overview, crafting plain-language summaries for the general public. She collaborates with community artists to illustrate data flows and safeguards. For the Community Insight tier, she hosts listening sessions to surface local concerns and tailors disclosures—translated materials, accessibility options, and contextual explanations. The Technical Detail tier provides documentation for researchers, auditors, and partner organisations, referencing AEIP manifests, crosswalk tables, and persona obligations. Finally, the Secure Oversight tier grants regulators and oversight bodies access to sensitive information under strict controls, ensuring that transparency never compromises safety or privacy.

The appendix narrates how disclosures evolve during incidents. When a service experiences a partial outage, Rina updates each tier: immediate alerts for affected users, detailed incident reports for oversight bodies, and interpretive notes for the hermeneutic ledger. She references Appendices B, E, and M to align messaging with remediation, rights, and adversarial considerations. The story emphasises reciprocity: transparency invites feedback that shapes future disclosures.
}
\normative{
Transparency programmes SHALL implement tiered disclosures encompassing Civic Overview, Community Insight, Technical Detail, and Secure Oversight. Each tier MUST define audience, content scope, delivery channels, frequency, and safeguards. Disclosures SHALL reference relevant AEIP manifests, appendices, and persona obligations to maintain coherence across the canon.

Public disclosures SHALL be accessible, multilingual where necessary, and inclusive of disability accommodations. Secure Oversight disclosures MUST employ encryption, access controls, and audit logging aligned with Appendix I security requirements. Any restriction on disclosure SHALL be justified with documented risk assessments and reviewed periodically.

Transparency updates during incidents SHALL follow Appendix B timelines, ensuring timely communication while protecting sensitive data. Feedback collected through transparency channels SHALL be recorded in Appendix L interpretive notes and Appendix C change logs when disclosures lead to policy updates. Failure to maintain transparency tiers SHALL trigger meta-audit review and potential custodial corrective actions.
}
%---------------------------------------------------------
%---------------------------------------------------------