Normative Concept
Glossary
Scope: Provides authoritative definitions for terms used in this specification. These definitions
SHALL be treated as binding where referenced in the normative text.
AI OSI Stack
The layered governance architecture for artificial-intelligence systems defined by this
specification. It separates institutional, epistemic, and technical duties into interoperable
layers with verifiable evidence flows.
AEIP (AI Epistemic Infrastructure Protocol)
Canonical transport for governance evidence across layers. Formal message unit: AEIP
Frame.
AEIP Frame
Atomic, signed evidence container including frame identifier, origin-layer tag, timestamp,
custodian identifier, evidence hash, and attestation signature (see Chapter 12).
Attestation Chain
Ordered sequence of AEIP Frames establishing provenance, verification, and time-bound
integrity for a given artifact or event.
Attribution Ledger
Tamper-evident record linking Reasoning Artifacts and governance events to originators,
models, datasets, and custodians (see Chapter 8).
Bias Mitigation Coverage (BMC)
Percentage of outcomes tested against declared bias metrics across relevant populations
(see Chapter 5).
Capability Domain
Thematic area scored in the Maturity Model (see Chapter 13); includes Institutional
62
63
Readiness, Technical Assurance, Ethical Integration, Civic Engagement, and Continuous
Improvement.
Change Control Record (CCR)
Immutable entry documenting a modification to a deployment instance, including rationale,
scope, and rollback reference (see Chapter 9).
Civic Charter
Ratified instrument conferring mandate to operate under this specification; includes
oversight structure, renewal cadence, and right of redress (see Chapter 3).
Civic Interface
Structured channel for public participation, feedback, and deliberation (see Chapter 11).
Custodian
Accountable role responsible for implementing controls, maintaining evidence, and re-
sponding to audits at a given layer.
Custodial Duty Indicator (CDI)
Metric family quantifying fulfillment of custodial responsibilities across layers.
Decommission Protocol
Procedure for secure retirement of systems, models, and data, including token revocation
and AEIP updates (see Chapter 9).
Deployment Instance
Unique operational instantiation of a model or system, identified by deployment ID and
configuration hash (see Chapter 9).
Differential Privacy
Privacy-preserving technique limiting individual disclosure risk; application SHALL be
risk-proportional and documented in CCMs (see Chapter 5).
Drift Detection Rate (DDR)
Frequency with which epistemic drift is detected and remediated within mandated time
frames (see Chapter 5).
EEP (Epistemic Exchange Protocol)
Standardized interface for exchanging Reasoning Artifacts, including signing, validation,
and ontology alignment (see Chapter 8).
Ethical Metrics
Quantitative indicators of interpretability, robustness, and harm potential tied to declared
Alignment Criteria (see Chapter 6).
Governance Disclosure Statement (GDS)
Periodic publication summarizing conformance, metrics, incident response, and corrective
actions (see Chapter 10).
Governance Transport Layer (GTL)
Institutional interface linking operational systems to AEIP evidence flow and maturity
measurement (see Chapter 13).
64 APPENDIX D. APPENDIX D – NORMATIVE CONCEPT GLOSSARY
Integrity Ledger Entry (ILE)
Atomic, immutable governance record produced by a control action or event; signed and
time-sealed.
Interpretive Envelope
Contextual metadata attached to Reasoning Artifacts indicating scope, confidence, and
validity interval (see Chapter 8).
Institutional Maturity Metric (IMM)
Composite maturity score derived from capability domains and supporting evidence (see
Chapter 13).
Incident Report (IR)
Structured record describing an anomalous or harmful event, with remediation status and
cross-layer impact (see Chapter 9).
Model Card
Authoritative documentation for a model including purpose, architecture, datasets, evalu-
ation, limitations, and risks (see Chapter 6).
Override Authority
Human role empowered to halt or constrain system operations under declared conditions
(see Chapter 7).
PCI (Provenance Completeness Index)
Ratio of datasets with complete provenance to total active datasets (see Chapter 5).
Provenance Registry
Append-only record linking data origin, transformations, consent, and lawful basis to
current validity state (see Chapter 5).
Reasoning Artifact
Structured inference or conclusion emitted for downstream decision-making, signed and
enveloped per EEP (see Chapter 8).
Reproducibility Package
Code, configuration, and documentation necessary to recompute training and evaluation
(see Chapter 6).
Service-Level Governance Indicator (SLGI)
Operational metric mapping reliability and safety thresholds to ethical obligations in
deployment (see Chapter 9).
Tamper-Evident Custody Log (TECL)
Asset-level record linking physical identifiers to cryptographic seals and audit events (see
Chapter 4).
Translation Governance Protocol (TGP)
Process ensuring that translations remain semantically faithful to the English canonical
text and cross-mapped to normative clauses