% SPDX-License-Identifier: CC-BY-SA-4.0

% Canonical Edition: AI OSI Stack v5.0
% Section: Part IV – Interpretive and Applied Canon
% Created: 2025-11-09
% Placement: After Layer 8 / Before Appendices
% Related Layers: 0–8
% Related Schemas: see /schemas/
\narrative{
Chapter 24 confronts the governance paradox: AI now assists in drafting the very rules that constrain it. The narrative follows a multi-institution drafting sprint where human stewards, machine collaborators, and ledger-backed observers co-author policy updates. It documents recursive authorship loops, consent-based prompt sharing, and verification rituals that ensure synthetic contributions remain attributable and contestable. Transparency requirements are stress-tested as authors publish real-time provenance, diff trails, and rationale bundles.

Workshops explore how recursion alters power dynamics. Stewards delineate decision boundaries, ensure human veto authority, and maintain interpretive logs referencing Appendices L and O. The chapter closes with a constitutional ceremony where the co-authored governance package is ratified, accompanied by public attestations explaining every machine-assisted sentence. The paradox resolves through radical transparency: AI participation is acknowledged, bounded, and democratically accountable.
}

\subsection*{Triple Register}
\textbf{Narrative Intent:} The governance paradox addresses the fear that AI oversight becomes performative, outlining how to balance automation assistance with human accountability in rulemaking.
\textbf{Normative Clauses:}
\begin{itemize}
\item Policymaking teams \shall{} disclose algorithmic drafting inputs via \texttt{schemas/governance/ai-assisted-drafting.jsonld} for every major revision.
\item Oversight bodies \shall{} record paradox resolutions in \texttt{schemas/oversight-audit-memo.jsonld} so contradictions are publicly examined.
\item Custodians \should{} log final settlements within \texttt{schemas/integrity-ledger-entry.jsonld} to demonstrate how authority was ultimately exercised.
\end{itemize}
\textbf{Plain-Speak Summary:} This chapter makes sure AI help does not replace accountable governance. It documents where machines assist, how conflicts are reviewed, and who signs off. Readers can track the debate from draft to decision. That transparency keeps power grounded in human responsibility.

\normative{
Governance drafting processes \shall{} disclose machine assistance, including prompt archives, persona manifests, and provenance hashes linked to Appendices N and O. Custodians MUST maintain AI-assisted drafting schemas that capture authorship roles, review checkpoints, and human ratification events. Recursively generated text \shall{} remain subject to civic challenge, with expedited amendment pathways when communities contest outcomes.

Institutions leveraging AI in governance creation MUST uphold human veto power, document interpretive rationales, and preserve audit-ready transcripts. Public ledgers \shall{} publish summary attestations describing machine contributions, ensuring governance remains traceable, contestable, and anchored in human accountability.
}
% Rationale: Anchor Canon operational linkage to AEIP evidence pathways for Phase 1+ synthesis.
% [SYNTHESIZED v5 PH1+]
\begin{quote}
\textit{Operational Note. Canon 24 links to AEIP decision rationale risk_acceptance entries and Layer~6 deployment dossiers, translating philosophical guardrails into measurable override thresholds.}
\end{quote}

\n\subsection*{Verification and Enforcement}
Conformance is evidenced through artefacts \texttt{schemas/governance/ai-assisted-drafting.jsonld}, \texttt{schemas/oversight-audit-memo.jsonld}, and \texttt{schemas/integrity-ledger-entry.jsonld} and corresponding AEIP audit records.
