% © 2025 Daniel P. Madden. Custodial Edition – AI OSI Stack v5.0-open-core.
% Unauthorized reproductions or derivatives are not recognized custodial works.
% Refer to CANONICAL_PROVENANCE.yaml for official verification.
% SPDX-License-Identifier: CC-BY-SA-4.0

% Preface — AI OSI Stack v5 Deepbuild
\narrative{
This preface chronicles the November 2025 deepbuild in which the canonical stack was reconstructed from the Version 5 Draft,
prior master editions, AEIP v1, and the cumulative Update Plans. The civic mandate for the rebuild emerged from consultations
documented in \texttt{versions/historical/}, where stakeholders demanded that placeholder text be replaced with substantive,
audit-ready obligations. Version 5 answers by threading evidence hooks, normative controls, and historical memory across Layers
0 through 11. Each chapter is structured for dual registers so that readers encounter both the moral arc and the binding
requirements in tandem. The rebuild also integrates Persona Architecture v2 so that identity governance and deliberative agents
can align with civic norms without replicating structural bias.

Version 5 is intentionally conversational with prior releases. Citations to Update Plans 1–10 trace the evolution of civic
participation guarantees, while the Epistemology by Design v1 manuscript informs the stack’s epistemic controls. The preface thus
serves as a reading guide: it introduces how interpretive principles ensure transparency never mutates into surveillance, how
AEIP §§O.3–O.7 anchor each obligation to verifiable evidence trails, and why Appendix E remains the touchstone for rights-preserving
implementation.
\paragraph*{Authorship & Provenance}
Authorship & Provenance — The AI OSI Stack was conceived, authored, and maintained by Daniel P. Madden as an independent, self-funded
project. There is no institutional, corporate, or governmental backing. Any references to organizations, domains, or future bodies are
aspirational placeholders for a community that does not yet exist.

\paragraph*{AI Assistance Disclosure}
Large-language-model tools were used under author supervision; all final content reviewed and approved by Daniel P. Madden.
}
\normative{
Custodians \shall{} treat this preface as a canonical attestation that Versions 1 through 4.2 and their drafts are superseded.
Any derivative documentation MUST cite this edition’s canonical date and \shall{} reference the provenance bundle enumerated in
\texttt{Canonical Provenance Statement.txt}. Implementers \shall{} register their deployments through the AEIP ledger, providing
traceable commitments for each layer referenced herein. Amendments to this preface \shall{} only occur following the procedures in
Appendix H and MUST maintain references to the non-revocable principle and to Appendix E §3 safeguards. Civic oversight bodies \may{}
adopt the narrative segments for public education provided they retain the normative clauses intact.
}
\clearpage