% AI OSI Stack v5 — Canonical Edition
% Daniel P. Madden — Independent AI Researcher
% Created: 2025-11-04
% File: 03_Preface.tex
% Purpose: Dual-register preface with supersession notice and custodial continuity
\narrative{
Version 5 emerges after a four-edition arc that began as a network governance experiment and matured into a civic infrastructure blueprint. Versions 1–3 established the vocabulary of layers and the promise that algorithmic systems could be governed like public utilities. Version 4 iterated through annexes, update plans, and public commentaries that refined the AEIP evidence backbone and introduced federated custodianship. Version 5 now consolidates those learnings into a canonical release authored between September and November 2025, ensuring that every chapter carries forward the commitments negotiated with communities, regulators, and implementers. Custodial continuity is preserved through a maintained provenance chain: the ledger of draft hashes, signatures, and crosswalks archived on \texttt{danielpmadden.com} is referenced throughout this edition so that readers can verify lineage without ambiguity.

This preface is offered to orient citizens, policymakers, engineers, and legal stewards to the narrative handoff from past editions. Update Plans 1–10, the Canonical Provenance Statement, and the AEIP lifecycle validators have been harmonised to prevent institutional memory loss. The custodianship team maintains stewardship not merely as archival duty but as an active civic role that connects release management with community oversight forums. Continuity is also cultural: the narrative affirms that transparency remains a human-rights safeguard only when held in tension with privacy, autonomy, and appeal mechanisms. Readers are invited to approach this edition as both a living constitution and a readiness kit for verifiable AI governance.
}
\normative{
All prior releases from Version 1 through Version 4.2.x, including interim drafts and provisional supplements, ARE HEREBY SUPERSEDED as of 4 November 2025. The canonical text for governance, implementation, and citation purposes SHALL be the artifacts whose signatures and hashes are recorded within \texttt{INTEGRITY\_NOTICE.md} and mirrored in the custodial ledger described in Appendix C. Any custodian distributing this work SHALL confirm the manifest hash and signature bundle against the ledger prior to publication or deployment. Amendments or errata SHALL document lineage references, SHALL cite superseded identifiers, and SHALL register new hashes within the AEIP integrity ledger before claiming canonical status. The non-revocable principle “Transparency must never become surveillance.” SHALL remain unchanged across all derivative notices.
}
\clearpage
