% AI OSI Stack v5 — Canonical Edition
% Daniel P. Madden — Independent AI Researcher
% Created: 2025-11-04
% File: 05_Custodianship_Notice.tex
% Purpose: Dual-register custodianship notice with timeline and obligations
\narrative{
Custodianship for the AI OSI Stack v5 followed a sequenced release chain spanning September through November 2025. In early September, draft harmonisation began with Update Plan 8, mapping feedback from civic forums into revised Layer narratives. October introduced Update Plans 9 and 10, integrating persona architecture requirements and AEIP lifecycle validators while publishing interim hashes for community verification. Between 1–3 November, custodians rehearsed the canonical signing ceremony, cross-checking provenance artifacts against the public ledger on \texttt{danielpmadden.com}. On 4 November 2025 the canonical release was sealed, with mirrored signatures committed to the Git repository and notarised within \texttt{INTEGRITY\_NOTICE.md}. This notice keeps that timeline legible so that every reader understands how stewardship moved from draft to declaration.

Custodial continuity is maintained by a multidisciplinary team that includes the author, legal reviewers, civic observers, and technical maintainers. Their responsibilities extend beyond archival duty: they coordinate public briefings, maintain Annex crosswalks, and ensure that transparency never mutates into surveillance by enforcing privacy-preserving publication practices. Each checkpoint in the release chain includes dual attestation—one narrative, one normative—to affirm that the civic story and the binding obligations evolve together.
}
\normative{
Custodians SHALL log every release checkpoint, including draft hashes, reviewer acknowledgements, and signature bundles, within the canonical provenance ledger referenced in Appendix C and Appendix H §2. Only artifacts whose hashes appear in \texttt{INTEGRITY\_NOTICE.md} SHALL be represented as canonical; any divergence MUST be disclosed and SHALL NOT claim custodial authority. Succession or delegation of custodial duties SHALL be registered through AEIP custodial records, SHALL include public notice within seven days, and SHALL document continuity safeguards that honour the non-revocable principle. Failure to publish renewal attestations within twelve months of the last canonical release SHALL trigger an escalation review involving civic overseers identified in Appendix N §1.
}
\clearpage
